Dynamic Load Balancing for\\Heterogeneous Systems

### 2011-12-05 paper/paper.tex
Polish.

\begin{abstract}
Leveraging maximum performance from specialized hardware is one of the
major problems facing computer scientists today.  General-purpose computing
on graphics processing units (GPGPU) is a technique that is becoming
increasingly popular for handling computations typically performed by the
CPU.  However, many GPGPU algorithms place the entire workload on the Graphics
processing unit (GPU) leaving the CPU idle for the duration of the computation.
This paper investigates the effect various load-balancing schemes across the CPU
and GPU have on overall application performance.  Additionally, the discrete and
fused GPU architectures, and their impact on performance, will be investigated.
\end{abstract}

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
\IEEEPARstart{L}{everaging} the maximum performance from specialized hardware is one
of the major problems facing computer scientists today.  The unique architecture of the
GPU gives it a place as a massively parallel hardware accelerator.  Programming frameworks
such as OpenCL allow programmers to utilize the stream processors of the GPU for non-graphics
data.  However, because of the difficulty in programming for this hardware, most of the
effort is spent on trying to achieve performance from the GPU, while the CPU is left
unused.  By also making use of CPU to process a chunk of the data, we can achieve higher
utilization of resources and therefore increase overall program performance.

@@

We expect that by using GPU+CPU processing we will improve the performance of the
application, when compared to CPU-only or GPU-only.  That is, consider a workload of size
$n$, a GPU that requires $t_g$ time to perform one block of computation, and a CPU that
requires $t_c$ time to perform one block of computation. We expect that the GPU would
require time $nt_g$ to perform a given computation, while the CPU would require $nt_c$
time to perform the same computation.  Optimally, we would expect the load-balancing
scheme to require the following time to perform the computation.
\[\max{\{(n-m)t_g, mt_c}\} + t_d + t_o\]
Where $m$ is the amount of the workload given to the CPU ($m \leq n$), $t_d$ is the data
transfer cost, and $t_o$ is any overhead incurrent from synchronization costs of the load
balancing scheme or otherwise.  The above will be faster than just the GPU iff
$t_g - \max{\{(n-m)t_g, mt_c}\} < t_d + t_o$.

@@

In addition, the CPU typically has a deep pipeline to increase throughput of the system.
Longer and longer pipelines increase the utilization of every element found on the CPU,
further increasing latency-critical performance.  Performance of this pipeline is in danger
of being flushed on branching codes, but advanced branch prediction and speculative execution
are able to greatly reduce the amount of times where a pipeline flush must occur.  Out-of-order
execution of instructions in the program also improves performance for single threads on the
CPU.

@@

\subsection{GPU Architecture}
The GPU almost acts a foil to the CPU architecture.  Instead of optimizing the
architecture of the device to increase performance of a single thread, the GPU
architecture is specifically designed for large throughput performance.  Large
numbers of threads can be easily created an run on a GPU.  Large register files
enable tens of thousands of threads to be run on today's GPU architecture with
very little penalty for context switching.

@@

The memory architecture for the GPU is very different than that found on the CPU. 
Instead of a large hierarchy of caches to hide memory latency for single thread
performance the latency to memory is instead hidden through aggressive context
switching.  Newer GPU architectures may contain privative caching schemes.
Each SM on the GPU also contains a local memory memory space which can accessed 
at much lower latency .  This memory space is typically used as a user-managed
cache. The memory hierarchy for the GPU architecture is given by Figure~\ref{fig:gpu_arch}.

@@

\section{The OpenCL Framework}
For this project, we will use the OpenCL Framework for our kernel launching.  OpenCL
is an application framework for heterogeneous parallel computing.  It includes an
API for creating and transferring memory to acceleration devices, as well as dynamic
compilation of application kernels that are run on the accelerators.Using OpenCL 
allows us to write the host code only once, while allowing us to run the core 
application kernel on the CPU, GPU, or both.

In OpenCL each thread is logically mapped to a work-item.  These work-items are grouped into
workgroups.  These workgroups map to a SM on a discrete GPU.  For instance, these workgroups
share access to a SMs local memory and can perform a barrier with all the other work-items
in the workgroup.  In the case of a CPU accelerator, the workgroups still exist, but optimizations
such as manual caching via local memory will not be effective.

Devices are grouped into contexts, which contains memory buffers and so on.  A queuing system
is used by OpenCL to facilitate kernel execution.  Kernels are submitted to the queue along with
the size of each workgroup, as well as number of workgroups to execute.  This queue is also 
responsible for reading and writing memory buffers.  The queue is in-order, meaning that 
the queue will wait until previously enqueued items have finished before starting work 
on more recently queued items.

For our purposes, we decided to use a separate context for each device, as well as a separate
command queue for each context.  We decided to use a separate context for each device, as it would
allow for finer-grained control of data transfers and residency. Using a separate command queue
for each device will facilitate multiple devices to be used simultaneously, as each command queue
operates on its own separate thread.

@@

First, the reduction benchmark was run on the discrete GPU platform with a variety of
different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
due to the GPU not utilizing all of its stream processors during a reduction operation, as
discussed previously.  It is also the case that the large amount of synchronization that must
occur in each iteration of the GPU reduction, and the number of kernel launches required for
each iteration that is not present in the CPU reduction. In addition, because the CPU had better
performance on the kernel alone, it saw an even greater improvement when the data transfer costs
were considered, with the CPU-only case requiring no data movement at all. Unfortunately, none of
the static load balancing schemes were able to outperform the CPU-only scheme; all the static times were
somewhere in between the CPU-only and GPU-only times.  Additionally, the dynamic load balancing
scheme performed significantly worse than all the other schemes, and more than 100\%
worse than just the CPU.  The cause for this degradation in performance was likely due to the
overhead of scheduling threads and communication between those threads, as well as moving data to
the GPU in small chunks when the GPU showed very poor performance compared to the CPU.

@@

Next, consider Figures \ref{fig:reduce_fused}, \ref{fig:vector_fused}, and
\ref{fig:vector_plus_fused}. The figures demonstrate a variety of different
load balancing options for the Reduction, VectorAdd, and VectorAdd+
benchmarks, across data sizes ranging between 1~MB and 16~MB on a Fused CPU+GPU platform.
Each figure depicts the total execution time for the CPU-only and GPU-only computations,
as well as static load balancing with 25\%, 50\%, and 75\% of the data on the
GPU.  In each of the figures, it's evident that the GPU performs significantly
faster than the CPU for the given benchmarks, with the static partitioning
schemes lying somewhere between the two.  As you can see in the figures, while the
overall performance numbers have changed between the discrete and fused architectures
the major trends from the previous graphs still hold true.

@@

Finally, consider Figure \ref{fig:vector_static_fused}, where the overall runtime
for the static load balancer is compared against both the CPU-only and GPU-only
runtimes for 0.1\%, 1\%, 10\%, 25\%, 50\%, 75\%, 99\%, and 99.9\% static data
ratios for the 16~MB vector add benchmark.  Figures \ref{fig:reduce_static_fused}
and \ref{fig:vector_plus_static_fused} show similar results for the reduction and
vector add+ benchmarks.  In these benchmarks, the downward trend as more data is
executed on the GPU becomes more evident.  What's also interesting is that the
static load balancer actually performs worse than the CPU-only case when only a
small amount of the data ($\leq$1\%) is executed on the CPU in the reduction and
vector add benchmarks.

@@

>>>>

\begin{abstract}
Leveraging maximum performance from specialized hardware is one of the
major problems facing computer scientists today.  General-purpose computing
on Graphics Processing Units (GPGPU) is a technique that is becoming
increasingly popular for handling computations typically performed by the
CPU.  However, many GPGPU algorithms place the entire workload on the Graphics
Processing Unit (GPU) leaving the CPU idle for the duration of the computation.
This paper investigates the effect various load-balancing schemes across the CPU
and GPU have on overall application performance.  Additionally, the discrete and
fused GPU architectures, and their impact on performance, has been investigated.
\end{abstract}

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
\IEEEPARstart{L}{everaging} the maximum performance from specialized hardware is one
of the major problems facing computer scientists today.  The unique architecture of the
GPU gives it a place as a massively parallel hardware accelerator.  Programming frameworks
such as OpenCL and CUDA allow programmers to utilize the stream processors of the GPU for non-graphics
applications.  However, because of the difficulty in programming for this hardware, most of the
attention is spent on trying to achieve optimal performance on the GPU to leverage the unique hardware
capabilites, while the CPU is left
unused.  By also making use of CPU to process a portion of the overall workload, we are able to achieve higher
utilization of resources and therefore increase overall program performance.

@@

We expect that by using GPU+CPU processing we will improve the performance of the
application, when compared to CPU-only or GPU-only.  That is, consider a workload of size
$n$, a GPU that requires $t_g$ time to perform one block of computation, and a CPU that
requires $t_c$ time to perform one block of computation. We expect that the GPU would
require time $nt_g$ to perform a given computation, while the CPU would require $nt_c$
time to perform the same computation.  Optimally, we would expect the load-balancing
scheme to require the following time to perform the computation.
\[\max{\{(n-m)t_g, mt_c}\} + t_d + t_o\]
Where $m$ is the amount of the workload given to the CPU ($m \leq n$), $t_d$ is the data
transfer cost, and $t_o$ is any overhead incurrent from synchronization costs of the load
balancing scheme or otherwise.  Note that there is an obvious minimum for this function the
point where the GPU and CPU workloads are equal.  The above will be faster than just the GPU iff
$t_g - \max{\{(n-m)t_g, mt_c}\} < t_d + t_o$.

@@

In addition, the CPU typically has a deep pipeline to increase throughput of the system.
Longer and longer pipelines increase the utilization of every element found on the CPU,
further increasing latency-critical performance.  Performance of this pipeline is in danger
of being flushed on branching codes, but advanced branch prediction and speculative execution
are able to greatly reduce the amount of times when a pipeline flush must occur.  Out-of-order
execution of instructions in the program also improves performance for single threads on the
CPU.

@@

\subsection{GPU Architecture}
The GPU almost acts a foil to the CPU architecture.  Instead of optimizing the
architecture of the device to increase performance of a single thread, the GPU
architecture is specifically designed for high-throughput performance.  Large
numbers of threads can be easily created an run on a GPU.  Large register files
enable tens of thousands of threads to be run on today's GPU architecture with
very little penalty for context switching.

@@

The memory architecture for the GPU is very different than that found on the CPU. 
Instead of a large hierarchy of caches to hide memory latency for single thread
performance the latency to memory is instead hidden through aggressive context
switching.  Newer GPU architectures may contain privative caching schemes, but the majority
of GPUs still on the commodity graphics market do not have this feature.
Each SM on the GPU also contains a local memory memory space which can accessed 
at much lower latency .  This memory space is typically used as a user-managed
cache. The memory hierarchy for the GPU architecture is given by Figure~\ref{fig:gpu_arch}.

@@

\section{The OpenCL Framework}
For this project, we will use the OpenCL Framework for our kernel launching.  OpenCL
is an application framework for heterogeneous parallel computing.  It includes an
API for creating and transferring memory to acceleration devices, as well as dynamic
compilation of application kernels that are run on the accelerators. Using OpenCL 
allows us to write the host code only once, while allowing us to run the core 
application kernel on the CPU, GPU, or both.

In OpenCL each thread is logically mapped to a so-called ``work-item''.  These work-items are grouped into
workgroups.  These workgroups map to a SM on a discrete GPU.  For instance, these workgroups
share access to a SMs local memory and can perform a barrier with all the other work-items
in the workgroup.  In the case of a CPU accelerator, the workgroups still exist, but optimizations
such as manual caching via local memory will not be effective.

Devices are grouped into contexts, which contains memory buffers and device specific information.  A queuing system
is used by OpenCL to facilitate kernel execution.  Kernels are submitted to the queue along with
the size of each workgroup, as well as number of workgroups to execute.  This queue is also 
responsible for reading and writing memory buffers.  The queue is in-order, meaning that 
the queue will wait until previously enqueued items have finished before starting work 
on more recently queued items.

For our purposes, we decided to use a separate context for each device, as well as a separate
command queue for each context.  Using a separate command queue
for each device will facilitate multiple devices to be used simultaneously, as each command queue
operates on its own separate thread. We also decided to use a separate context for each device, as it would
allow for finer-grained control of data transfers and residency.

@@

\subsection{Discrete GPU Performance}
First, the reduction benchmark was run on the discrete GPU platform with a variety of
different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
due to the GPU not utilizing all of its stream processors during a reduction operation, as
discussed previously.  It is also the case that the large amount of synchronization that must
occur in each iteration of the GPU reduction, and the number of kernel launches required for
each iteration that is not present in the CPU reduction. In addition, because the CPU had better
performance on the kernel alone, it saw an even greater improvement when the data transfer costs
were considered, with the CPU-only case requiring no data movement at all. Unfortunately, none of
the static load balancing schemes were able to outperform the CPU-only scheme; all the static times were
somewhere in between the CPU-only and GPU-only times.  Additionally, the dynamic load balancing
scheme performed significantly worse than all the other schemes, and more than 100\%
worse than just the CPU.  The cause for this degradation in performance was likely due to the
overhead of scheduling threads and communication between those threads, as well as moving data to
the GPU in small chunks when the GPU showed very poor performance compared to the CPU.

@@

\subsection{Fused APU Performance}
Next, consider Figures \ref{fig:reduce_fused}, \ref{fig:vector_fused}, and
\ref{fig:vector_plus_fused}. The figures demonstrate a variety of different
load balancing options for the Reduction, VectorAdd, and VectorAdd+
benchmarks, across data sizes ranging between 1~MB and 16~MB on a Fused CPU+GPU platform.
Each figure depicts the total execution time for the CPU-only and GPU-only computations,
as well as static load balancing with 25\%, 50\%, and 75\% of the data on the
GPU.  In each of the figures, it's evident that the GPU performs significantly
faster than the CPU for the given benchmarks, with the static partitioning
schemes lying somewhere between the two.  As you can see in the figures, while the
overall performance numbers have changed between the discrete and fused architectures
the major trends from the previous graphs still hold true.

@@

\subsection{Load Balancing Performance}
Consider Figure \ref{fig:vector_static_fused}, where the overall runtime
for the static load balancer is compared against both the CPU-only and GPU-only
runtimes for 0.1\%, 1\%, 10\%, 25\%, 50\%, 75\%, 99\%, and 99.9\% static data
ratios for the 16~MB vector add benchmark.  Figures \ref{fig:reduce_static_fused}
and \ref{fig:vector_plus_static_fused} show similar results for the reduction and
vector add+ benchmarks.  In these benchmarks, the downward trend as more data is
executed on the GPU becomes more evident.  What's also interesting is that the
static load balancer actually performs worse than the CPU-only case when only a
small amount of the data ($\leq$1\%) is executed on the CPU in the reduction and
vector add benchmarks.

@@

Additionally, the use of only applications which have a static workload per thread limit the
effectiveness of the dynamic scheduler.  The ability to a priori know the thread workload makes
the static load-balancing scheme more effective than the dynamic scheme.
Using a kernel with a varied workload per thread
would have made it possible for the dynamic scheduler to see improved performance when
compared to the static mapping of data.
